---
title: "Notebook support: IA640 Dashboard lab (Pt 2)"
author: "Joe Skufca"
date: "23 November, 2019"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
    theme: spacelab
    highlight: tango
    toc_float: true
    collapsed: false
editor_options: 
  chunk_output_type: inline
---
## Interactivity

This notebook provides *support* to the IA640 lab on building dashboards.  I will use this notebook to build the coded driven *elements* of the dashboard, and then cut and paste to insert to file that builds the dashboard.

### Packages

We build from some of the visualization we used in part 1, but add a bit of interactivity.



I am not building the dashboard here, so I simply am loading the standard tools I use for analysis.

```{r collapse=TRUE}
library(knitr) # to create nice documents in R
library(tidyverse) # loads ggplot2, dplyr,tidyr,readr,purr,tibble
library(forcats) # working with categorical variables
```

As we will be working with dates, I also load the `lubridate` package (to make it easy to work with date-time information).

```{r}
library(lubridate)
```


We will need a couple more, but I will hold off on loading so that I can show exactly where they come into play.

### Load some data

We continue with the OpenDataPhilly website data, https://www.opendataphilly.org/ .  These particular data are related to complaints filed against the police.

> You should make sure you have downloaded the files from moodle and put them in your working directory.

The three files give documentation about the complaints, the demographics of the complainant, and demographic information about the police involved (along with and disciplinary action taken).

```{r message = F}
df1=read_csv("ppd_complaints.csv")
df2=read_csv("ppd_complainant_demographics.csv")
df3=read_csv("ppd_complaint_disciplines.csv")
```

### Some Dashboard components


##### Monthly counts of complaints.

Perhaps we might be interested a monthly count of complaints.   We can use the `floor_data()` function from the lubridate package to help aggregate that data.

Note that I have added a bit beyond the default ggplot call to get nice labels on the x-axis.

```{r}
dfA=df1 %>% group_by(week_reps=floor_date(date_received, "week")) %>%  summarise(reports=n()) 

dfA %>% filter(week_reps >= "2019-09-01") %>%
 ggplot(aes(x=week_reps,y=reports)) + geom_col() +
  scale_x_date() +
  ggtitle("Complaints filed by month (year to date)")+
  theme_minimal()
```

##### Disciplinary actions taken

Note that in `df3` , if the investigation found no wrongdoing ("No Sustainted Findings") then no disciplinary action was taken ("Not Applicable").  But of those where disciplinary action was taken, what was the outcome.  


```{r}
dfB =df3 %>% filter(disciplinary_findings != "Not Applicable")

dfB %>% 
  ggplot(aes(x=disciplinary_findings)) + 
  geom_bar() 
```

Suppose we want to break this appart by year of the complaint.  Then we have to do a little work, because dates are available only in the `df1` dataframe.  So we will have to join dataframes, and repeat the above steps, but adding faceting.

```{r}
left_join(df3,df1,by="complaint_id") %>% 
  filter(disciplinary_findings != "Not Applicable") %>%
  ggplot(aes(x=disciplinary_findings)) + 
  geom_bar(aes(fill=disciplinary_findings)) +
  facet_wrap(~year(date_received),ncol = 5)+
  theme_minimal()+
  theme(axis.text.x=element_blank()) # get rid of the text labels on the x axis

  
```

##### Complaints filed by district

Lets just get a table of data, rather than a plot, for the current month.


```{r}
df1 %>% 
  filter(date_received>"2019-09-01") %>%
  group_by(district_occurrence) %>%
  summarise(complaints=n()) %>% arrange(complaints)
```

```{r}
df1 %>% 
  filter(date_received>"2019-09-01") %>%
  count(district_occurrence,sort=TRUE) 
```

##### Perhaps we are concerned about the demographics of the accused officers

Let's construct a two-way table of police officer's involved in these incidents.

We will take advantage of the `tabyl` command from the `janitor` package and pretty things up with `kable`.

```{r}
library(janitor)
df3 %>% tabyl(po_race,po_sex) %>%  kable()
```

##### Demographics on complainants

```{r}
df2 %>% tabyl(complainant_race,complainant_sex) %>%  kable()
```

##### Let's focus on current month

```{r}


df1 %>% filter(date_received >= "2019-09-01") %>%
ggplot(aes(x=date_received)) + geom_bar() +
  scale_x_date() +
  ggtitle("Complaints filed by month (year to date)")+
  theme_minimal()
```

##### Interaction

Let's add some capability for the plots to be interactive.  I will add something using the `dygraphs` package, looking at reports on a weekly basis.



```{r}
library(dygraphs)
library(tsbox)
```

First, we create a dataframe with weekly reports.

```{r}
dfA=df1 %>% group_by(week_reps=floor_date(date_received, "week")) %>%  summarise(reports=n()) 
```

Then plot, along with a control for zooming.

```{r}
dfA  %>% ts_xts() %>% 
  dygraph() %>% dyRangeSelector() %>% 
  dyOptions(stepPlot = TRUE)
```

Usually, I make a graph that captures the data, then I work on prettying up that graph.  Maybe I look online for a tutorial, but lots of places could be of use, to include the "help" on the command.



```{r}
dfA  %>% ts_xts() %>% 
  dygraph(ylab="Number of complaints filed") %>% dyRangeSelector() %>% 
  dyOptions(stepPlot = TRUE)
```

##### Detailed (and interactive) look at this weeks complaints, by district

The `plotly` set of tools has its own plotting capability, but it also allows us to take a ggplot created graph and make it interactive.

```{r}
library(plotly)
```

I will recreate one of the graphs from above, simply by piping the result into a `plotly` plotting command.


```{r}
# create the plot
p1=df1 %>% filter(date_received >= "2019-09-01") %>%
ggplot(aes(x=date_received)) + 
         geom_bar() +
  scale_x_date() +
  ggtitle("Complaints filed by month (year to date)")+
  theme_minimal() 
  


ggplotly(p1)
```

In this case, a little extra work is required to get the date formatted correctly on the tooltip.

```{r}
# create the plot
p1=df1 %>% filter(date_received >= "2019-09-01") %>%
ggplot(aes(x=date_received,text = paste('Date: ', as.Date(date_received)))) + 
         geom_bar() +
  scale_x_date() +
  ggtitle("Complaints filed by month (year to date)")+
  theme_minimal() 
  
ggplotly(p1,tooltip = c("y","text"))
```

But what I really wanted was to do a stacked barchart by type of complaint.

```{r}
# create the plot
p1=df1 %>% filter(date_received >= "2019-09-01") %>%
ggplot(aes(x=date_received,text = paste('Date: ', as.Date(date_received)))) + 
         geom_bar(aes(fill=general_cap_classification)) + 
  labs(fill="Classsification of complaint")+
  scale_x_date() + coord_flip()+
  ggtitle("Complaints filed by month (year to date)")+
  theme_minimal() 
  
ggplotly(p1,tooltip = c("y","text"))
```
