---
title: "ggplot2 with the cereals dataset"
author: "Joe Skufca"
date: "October 2019"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
    theme: spacelab
    highlight: tango
    toc_float: true
    collapsed: false
editor_options: 
  chunk_output_type: inline
---


My intention here is not to provide a tutorial on ggplot2.  Those already exist.  What I did want to show you is how I (conceptually) might have manipulated the dataset we were using for the cereals homework assignments to make a few plots.

# Load packages

The `tidyverse` should be sufficient.

```{r}
library(tidyverse)

```

# Get the data

```{r}
df1=read_csv('cereals.csv') # let function guess the column types
```



### Why ggplot2?

`ggplot2` is a bit more complicated for simple plotting, but for more general analysis of a dataset, ggplot2 provides a robust set of static visualizations.  *Additionally*, it is built on a conceptual framework, **The Grammar of Graphics**, that matches closely with the theoretical underpinnings of data visualization that we are considering in this course.


> Note: In this session, we won't be going after "graphic excellence."  Rather, we are focused on connect the theoretical framework of *Visualization Analysis and Design* with the visualization implementation using ggplot2 and related packages.



# A first graph

To make any graph, we need data (our loaded dataframe).  The *what* component of the abstraction.  

For the task abstraction (the *why*) let's leave that loose, and simply suppose that we want to compare (what we think) are key nutritional variables and see if there is a relationship.

### Are sugar and calories related?

`Sugars` and `Calories` are both quantitative variables in our dataset, and one way to look at the relationship is via a scatterplot.

```{r}
df1 %>% ggplot(aes(x=Sugars,y=Calories))+
  geom_point()
```

"aes":  aes - aesthetics - assigns a variable to a particular channel.  We are telling the graphics command to use horizontal position (x) to encode the `Sugars` value and vertical position (y) to encode `Calories`.

"geom_point()": tells the system what type of *mark* to make on the page. (In this case, a point.)

##### **(1)** Edit the code so that it (instead) plots Sugars along the horizontal, but plots the `Carbohydrates` on the vertical.


```{r}
# Edit me
df1 %>% ggplot(aes(x=Sugars,y=Calories))+
  geom_point()
```


### The data

Let's take a quick look at our data:

```{r}
head(df1)
```

### Let's use more aesthetics.

In Munzner's terminology, we there are a number of ways to encode a variable (called channels).  In ggplot2, the `aes` is where we declare the particular association of channel and variable that we want to use.

We will continue with the scatter plot (geom_point()), which, according the the help page, understands the following aesthetics:


* x
* y
* alpha
* colour
* fill
* group
* shape
* size
* stroke

Let's return to Sugars and Calories to play a bit more with some of these.


```{r}
df1 %>% ggplot(aes(x=Sugars,y=Calories))+
  geom_point(aes(colour=Sugars,size=Calories,fill=Calories),
             shape="circle filled",stroke=3)
```




#  Manipulation of that plot by adding layers

In ggplot2, a created plot is an object, and can be stored as a variable.

```{r}
p= df1 %>% ggplot(aes(x=Sugars,y=Calories))+
  geom_point()
```


We can then add *layers* to that plot.



#### Adding a smooth curve (based on a statistical model)

```{r}
p=p+geom_smooth()
p
```
####  Manipulation of that plot by adding a theme

Adding a theme manipulates broad characteristics

```{r}
p=p+theme_bw()
p
```

#### Multiple layers can be added all at once

Here I also added a jitter effect

```{r}
p=ggplot(df1,aes(x=Sugars,y=Calories))+
  geom_point()+theme_bw()+geom_jitter(width=.3,height=.3)
p
```

#### Using transparency to deal with overplotting

An alternative way to deal with overplotting it to use transparency - the alpha channel

```{r}
p=df1 %>% ggplot(aes(x=Sugars,y=Calories))+
  geom_point(alpha=.2,size=3)+theme_bw()
p
```



# aes - aesthetics - assigns a variable to a particular channel

Assign the color channel to "Shelf".

```{r}

```
Note that "Shelf" is a numerica variable, so the colorscale is continuous

# aes - aesthetics - assigns a variable to a particular channel

Lets treat Shelf as categorical

```{r}
p=df1 %>% mutate(Shelf=factor(Shelf)) %>% ggplot(aes(x=Sugars,y=Calories,color=Shelf))+
  geom_point()+theme_bw()+geom_jitter(width=.3,height=.3)
p
```


# geom - geometry - selects a mark, but (sometimes) may also identify an idiom


Manufacturer

```{r}
p=df1 %>% mutate(Shelf=factor(Shelf)) %>% ggplot(aes(x=Sugars,y=Calories,color=Shelf))+
  theme_bw()+
  geom_text(aes(label = Manufacturer), position=position_jitter(width=.2,height=.2))
p
```

# Barcharts - a different geometry
\small

```{r}
p=df1 %>% ggplot(aes(x=Shelf))+
  theme_bw()+
  geom_bar()
p
```

# Stacked Barcharts 

Use the "fill" channel for the variable "Manufacturer."

```{r}

```

# The scatterplot matrix - because I asked you to
\small
Package GGally adds this functionality

```{r}
library(GGally)
df1 %>% select(Calories,Protein,Fiber,Carbohydrates) %>% ggpairs()
```

## A second data set

Let's play with MTCARS

```{r}
df2=mtcars
knitr::kable(df2)
```


```{r}
library(DT)
datatable(df2)
```


