---
title: "Demo on Dept budget number"
author: "Joe Skufca"
date:  "22 SEP 2020"
output: html_notebook
---

## Prepare Workspace

Load the packages we intend use.  I always load the tidyverse, along with whatever else I want.


```{r}
library(extrafont)
library(ggthemes)
library(tidyverse)
library(janitor)
library(readxl)
library(scales)
```


## Import data

I want to bring in the data from the dept-sales performance Excel.

```{r}
a2_DeptYearlyPerf <- read_excel("a2-DeptYearlyPerf.xls")
```

## Tidy the data

Transform first into long data.  I will use pivot_long from tidyr.

```{r}
df2 = a2_DeptYearlyPerf %>% 
  pivot_longer(c(2:9),names_to = c("qtr","budget_or_actual"),
               names_sep = " ",values_to = "amount") %>% clean_names()
```



## Transforms

I may need to work with the difference between budget and actual.


```{r}
df3= df2 %>% pivot_wider(names_from = budget_or_actual,
                         values_from = amount) %>% clean_names()
```

Lets compute a budget deviation column:

```{r}
df3 =df3 %>% mutate(budget_diff=budget-actual)
```


## Build an initial visualization


```{r}
df2 %>% ggplot(aes(x=dept,y=amount)) +
  geom_bar(aes(fill=budget_or_actual), stat = "Identity",position = "dodge")

```


Using Faceting


```{r}
df2 %>% ggplot(aes(x=qtr,y=amount)) +
  geom_bar(aes(fill=budget_or_actual  ), stat = "Identity",position = "dodge2") +
  facet_wrap(~dept,ncol=4)
```

But maybe I want to emphasize something else, like the budget difference.


```{r fig.asp=.4}
df3 %>% mutate(dept=as_factor(dept) %>% fct_reorder(-actual)) %>%
               ggplot(aes(x=qtr,y=actual/1000)) +
  geom_crossbar(aes(ymin=budget/1000,ymax=actual/1000,fill=(actual>budget)  ),color=NA) +
  geom_crossbar(aes(ymin=actual/1000,ymax=actual/1000  ),size=.3) +
  facet_wrap(~dept,ncol=4)+theme_minimal(base_size = 10,base_family = "Gill Sans MT")+
  scale_fill_manual(name="Comparison to \n Budget",
                          labels=c("Under \nBudget", "Over \nBudget"),
                    values = c("#5ab4ac","#d8b365"))+
  #theme(text = element_text(size = 10,family="GillSans")) +
  xlab("Quarter")+ylab("Expenditures ($k)")+
  ggtitle("Expenditures by department and quarter, with deviation from budget")+
  scale_y_continuous(breaks = breaks_width(200),limits = c(0,1000))
```

Try 2:

```{r fig.asp=.4}
df3 %>% mutate(dept=as_factor(dept) %>% fct_reorder(-actual)) %>%
               ggplot(aes(x=qtr,y=actual/1000)) +
  geom_crossbar(aes(ymin=budget/1000,ymax=actual/1000,fill=(actual>budget)  ),color=NA) +
  geom_crossbar(aes(ymin=actual/1000,ymax=actual/1000  ),size=.3) +
  facet_wrap(~dept,ncol=4)+
  theme_tufte(base_size = 10,base_family = "Gill Sans MT")+
  scale_fill_manual(name="",
                          labels=c("Under \nBudget", "Over \nBudget"),
                    values = c("#5ab4ac","#d8b365"))+
  xlab("Quarter")+ylab("Expenditures ($k)")+
  ggtitle("Expenditures by department and quarter, with deviation from budget")+
  scale_y_continuous(breaks = breaks_width(200),limits = c(0,1000))
```


## Reflection

Here's what I think ....

