---
title: "Sales Rep Graphs"
author: "Joe Skufca"
date: "Jun 14,2018"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( cache = T, fig.width=8)
```

Joel,

You requested that I help you grapple with how produce something like you were producing in Tableau.


```{r}
# load the package 
library(tidyverse)
library(readxl)
```

### Load the data


I generally find that Excel .xls files can be tricky, so I initially opened the data in excel and then saved as a .csv format.  In general, .csv , being a plain text readable format is much more portable, so I often do that first step outside of R.  

Having done so, I load the data in the standard way.

### Loading from .xls format

```{r}
af=read_xls("a2-SalesRepPerf.xls")
head(af)
```



## Tidying the data 

I will work from the `af` data frame.  



#### One minor fix before we start:
```{r}
names(af)[6]="Actual"   # Spaces in a variable name is bad syntax
names(af)[1]="Sales_rep"
```


#### Make the data frame "long"

As you are using `Actual` and `Quota`, they are a categorical variable.  

The original table is tidy, but you need it in the "long" format.

Additionally (although it won't be needed here) I will `select` only the relevant columns 


```{r}
afv=af %>% gather(key,val,c(2,6)) %>% select(Sales_rep,key,val)
afv
```

Here is the same thing, using the pivot command:

```{r}
afv2 = af %>% 
  pivot_longer(cols = c(Quota,Actual),names_to = "key",values_to = "val") #%>%
  #select(Sales_rep,key,val) 
afv2
```



### Using the long dataset

If we want to encode _Quota_ and _Actual_ information differently, then we need to use the `key` variable in the long data frame.


```{r}
ggplot(afv, aes(x=Sales_rep, y=val, fill = key))+
  geom_bar(position="dodge",stat = "identity")
```

### Or you could encode by facets 

```{r}
ggplot(afv, aes(x=key, y=val))+facet_wrap(~Sales_rep,ncol=13)+
  geom_col()+theme_bw()
```

### At this point, you can tweak the labels


#### Some minor tweeks

I want prettier scales (and I don't like the gray background)

```{r}
library(scales)
library(ggthemes)
ggplot(afv, aes(x=key, y=val))+facet_wrap(~Sales_rep,ncol=13)+
  geom_col()+
  scale_y_continuous(labels=dollar)+theme_tufte()
```


#### Perhaps a bit more work on the labels

Note - text on horizontal is easier to read.

```{r fig.height=8}
library(scales)
library(ggthemes)
ggplot(afv, aes(x=key, y=val))+
  facet_wrap(~Sales_rep,ncol=1,strip.position = "right",)+
  geom_col()+
  scale_y_continuous(labels=dollar)+theme_minimal() +
  coord_flip()+
  theme(strip.text.y = element_text(angle = 0))
```

## The Departmental data

## Load the data

```{r}
df2=read_excel("a2-DeptYearlyPerf.xls")
names(df2)[1]="Dept"
```

```{r}
df3=df2 %>% pivot_longer(c(2:9),
                         names_to = c("Qtr","Budget.or.Actual"),names_sep = " ",values_to = "value")
```

```{r}
df4= df3 %>% pivot_wider(names_from = Budget.or.Actual,values_from = value)
```

```{r}
df4 = df4 %>% mutate(Budget.difference=Budget-Actual)
```

```{r}
df3 %>% ggplot(aes(x=Dept,y=value)) + 
  geom_bar(aes(weight=Qtr,fill=Budget.or.Actual),stat = "Identity",position = "dodge")
```

```{r}
df3 %>% ggplot(aes(x=Dept,y=value)) + 
  geom_bar(aes(fill=Budget.or.Actual),stat = "Identity",position = "dodge")+
  facet_wrap(~Qtr)
```

```{r}
df3 %>% ggplot(aes(x=Qtr,y=value)) + 
  geom_bar(aes(fill=Budget.or.Actual),stat = "Identity",position = "dodge")+
  facet_wrap(~Dept)
```