---
title: "A Homework on Maps in R"
author: "Muhammad Rashid"
date: "11/15/2022"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 3
    theme: spacelab
    highlight: tango
    toc_float: yes
    collapsed: no
---
# Pedagogy Preliminaries

This material serves a dual purpose:

* It is the basis for a class lecture, for which I will work my way throught this document, focusing primarily on the connections to visualization.
* It provides a homework assignment framework, requesting that you do some coding to produce some of these interesting maps.

##### Citation

This material is based on a document created by Claudia Engle, available at https://cengel.github.io/rspatial/4_Mapping.nb.html .



#### To complete this assignment, follow these steps:

1. Download the `HomeworkMaps2020.Rmd` file from Moodle

2. Open `HomeworkMaps2020.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with your own name.

4. Supply your solutions to the homework by editing `HomeworkMaps.Rmd`.

5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `HomeworkMaps_YourNameHere.Rmd`, and submit both the `.Rmd` file and the `.html` output file on Blackboard.  (YourNameHere should be changed to your own name.)

##### Homework tips:

1. Recall the following useful RStudio hotkeys.

Keystroke | Description
------------|-------------------------------------------
`<tab>` | Autocompletes commands and filenames, and lists arguments for functions.
`<up>` | Cycles through previous commands in the console prompt
`<ctrl-up>` | Lists history of previous commands matching an unfinished one
`<ctrl-enter>` | Runs current line from source window to Console. Good for trying things out ideas from a source file.
`<ESC>` | Aborts an unfinished command and get out of the + prompt

**Note**: Shown above are the Windows/Linux keys.  For Mac OS X, the `<ctrl>` key should be substituted with the `<command>` (&#8984;) key.

2. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.

3. Run your code in the Console and Knit HTML frequently to check for errors.

4. You may find it easier to solve a problem by interacting only with the Console at first, or by creating a separate `.R` source file that contains only R code and no Markdown.

# Making Maps

Spatial data---creating, maintaining, analyzing, studying spacial processes---create an entire ecosystem of data issues that could easily fill multiple courses.  Our focus here is to focus very narrowly on certain **visualization tasks**, especially those most pertinent to a data analyst.  So ... there will be lots of things (from a spatial data perspective) that we simply brush upon here, but for which you could pursue further reading. 

### Packages

We will give preference to the "tidyverse" of packages related to "Tidy Data."    

First - some general tools that I *almost always* load:

Several spatial packages are load, where we will touch on only upon the elements we need for the visualizations.  R has a lot of packages related to geospatial analysis

```{r setup,collapse=TRUE}
## additional libraries needed for R code examples
library(sp)
library(sf)
library(rgdal)
library(classInt) 
library(RColorBrewer) # not a strictly a spatial library
library(viridis) #my current "go to" color scales
library(ggmap)
library(leaflet)
library(tmap)
#library(GISTools)
```


```{r collapse=TRUE}
library(knitr) # to create nice documents in R
library(tidyverse) # loads ggplot2, dplyr,tidyr,readr,purr,tibble
library(broom)  # because I find it useful
library(forcats) # working with categorical variables
```





R comes with a basic `plot` command, which can also be used for simple map viewing. In this workshop we will look into several alternatives to map spatial data with R. This list is of course not comprehensive, but should give you something to start with. For more packages see the "Visualisation" section of the [CRAN Task View](https://cran.r-project.org/web/views/Spatial.html), or simply take a look at that page to have a bit more appreciation for the spectrum of packages in R that are relevant to spatial analysis.

Of the packages mentioned  `spplot`  only takes `sp` objects, 
`tmap` and `leaflet` can also take `sf` objects. The [development version of `ggplot2`](https://github.com/tidyverse/ggplot2/releases) can take `sf` objects, though `ggmap` [seems to still have issues](https://github.com/tidyverse/ggplot2/issues/2130) with `sf`.


### Data needed:

* Shapefile of Philadelphia census tracts and homicide rates: [Philly3.zip](https://www.dropbox.com/s/pw4611p7x3h5sqa/Philly3.zip?dl=1) .  Inside of the zip file is a directory called `Philly3`.  

__Make sure that you place an unzipped copy of that directory as a subdirectory in your working directory.__

```{r}
## need this file for later
philly <- readOGR("Philly3", "Philly3", verbose = F)
names(philly)
```

Use `names()` to see the attributes that Claudia has added  to supplement the existing geospatial fields:

* _N_HOMIC_: Number of homicides (since 2006)
* _HOMIC_R_: homicide rate per 100,000 (Philadelphia Open Data)
* _PCT_COL_: % 25 years and older with college or higher degree[^1] (ACS 2006-2010)
* _mdHHnc_: estimated median household income (ACS 2006-2010)


#####  **(Q1)** What package is the `readOGR()` function from.

Inside Rstudio, drill down a little bit into `philly` so that you can appreciate the complexities of a spatial data.  In this instance, the geospatial objects are polygons - and defining the polygons is the majority of the data.  Additionally, this spatial dataframe includes data about each of the polygons.

> *rgdal* is an older package that focus on input and output - getting spatial data into R (from other standard sources) and producing files that could be read by other spatial software.


##  2. Choropleth mapping with `spplot` 


The *sp* package provides a standarized set of objects and methods for spatial analysis.  Many packages use the sp objects.

Base R has the ability to plot these objects.

```{r }
plot (philly)
```

`sp` comes with a plot command `spplot()`, which takes `Spatial*` objects to plot. `spplot()` is one of the earliest functions to plot geographic objects.

The `ssplot` command also allows for visualization, but you will (typically) want to select a data attribute to be plotted.

##### **(a)** Use the `spplot` command (make sure you installed and loaded `sp`) to make a chloropleth using Homicide Rate data.


```{r}
spplot(philly, 'HOMIC_R')
```

Based on what we have learned in this course, we might suspect that we can improve on the default coloring scheme.  Taking advantage the *viridis* color scale, and a continuous coloring scheme.

```{r }
spplot(philly,"HOMIC_R", col.regions = viridis(256))
```

We know the color scale has nice properties, but we still have this one area standing out with an extremely high homicide rate, which makes a large part of the map unreadable. So let's change the class intervals to quantiles. We will use `classIntervals` from the `classInt` library, something like:

        breaks_qt <- classIntervals(vector_of_values, 
                                    n = how_many_classes, 
                                    style = "quantile" [can be omitted -- the default])

    This returns an object of type `classIntervals`. Find out its structure. How would you access the break values?

```{r }
breaks_qt <- classIntervals(philly$HOMIC_R, n = 5,dataPrecision=2)
head(breaks_qt)
```

R is not always perfect.  We need to "dress up" this result a little bit, otherwise it will (sometimes) cut off part of the range of interest, which would leave some polygons without a color assigned.

```{r}
br <- breaks_qt$brks 
offs <- 0.0000001
br[1] <- 0 # because this data is lower bounded
br[length(br)] <- br[length(br)] + offs

```

Finally we use those breaks to set the `at=` argument in `spplot()`. Let's also set `main=` to add a title. 

```{r }
spplot(philly, "HOMIC_R", col.regions=viridis(length(br)), at = br,
       main = "Phil homicide rate per 100,000")
```

##### **(b)** In the above plot, I discretized the continuous variable into 6 levels.  Provide an "Infovis" based justification for this decision.

- **When we plot at first the map we were ploted it using the contiuous scale which clearly not showed the changes in colors on map and the visualization was not great. Some areas with extremly high homicide rate were colored such that map was unreadable. Therefore chosing intervals and discretizing the operation help us better visualize the map. as shown above using 6-intervals with breaks in between**

##### **(c)** Comment upon the good/bad of using *quantiles* to choose the breakpoint for discretization.

- **Quantiles classification carries same kind of features.Therefore, Linearly distributed responds well to quantile classsification.**

-  **Quantiles classification can result in misleading, unreadable outputs. We can group features into groups with distinct type of values **


___

Tweeking a little further, the biggest remainig issue is the legend, which shows as a graduated scale even though we have discretized the coloring. Here is how we can change this: 

    - Use the `cut()` function from the base package with the values from `philly$HOMIC_R` and the corrected breaks `br` to return a vector with the respective boundaries of the brackets. Use `?cut` if you need help.
    - Assign the output vector you get as a new column `HOMIC_R_bracket` to the `philly` attributes table. It will help to map the color based on the breaks. Take a look at the values. What object class is that vector?  
    - Remove the `at=` parameter in `spplot()` (which is only needed for continuous variables) and tell it to plot `HOMIC_R_bracket`.  

```{r }
philly$HOMIC_R_bracket <- cut(philly$HOMIC_R, br,include.lowest = T)
spplot(philly, "HOMIC_R_bracket", col.regions=viridis(5), main = "Philadelphia homicide rate per 100,000")
```


---

There are many more arguments for this function to provide additional plot parameters, like the legend position, labels, scales, etc. 

However, as you have seen, this can be quite tedious. 

As an alternative you may want to be aware of the [`GISTools`](https://CRAN.R-project.org/package=GISTools) package. It includes functions, like `choropleth()` to draw choropleth that are really just convenience functions that wrap around `spplot()`

## 3. Plotting simple features (`sf`) with `plot`

The `sf` package looks to connect R more closely with GIS industry standards and looks to be the follow-on to the `sp` package.  For a longer description, you can read https://cran.r-project.org/web/packages/sf/vignettes/sf1.html .

The `sf` package extends the base `plot` command, so it can be used on `sf` objects. If used without any arguments it will plot all the attributes, like `spplot` does.

To work in the *sf* package, we need an *sf object*.  We could do a conversion, (from the sp object we have been using), but I want to emphasize reproducibility.  

Let's go back, read from a file.  The reader command reads directly from the "shape" file, which is inside the Philly directory. 

```{r message = F}
philly_sf <-  st_read("Philly3/Philly3.shp")
plot(philly_sf)
```

Take a look at the structure of `philly_sf`.  You'll observe a more compact (simple) structure than from `philly`, where now the geometry is just one of the attributes.

To plot a fewer attributes we need to provide an object of class `sf`, which has been subsetted appropriately.  I am going to accomplish by using one of my *dplyr* tools (the `select()` function) using *chaining*:

```{r }
sf2= philly_sf %>% select(HOMIC_R,PCT_COL)
plot(sf2)
```

##### **(a)** Create your own plot of this sf object using the plot command, but where you plot 3 variables of your choosing.

```{r}
sf3= philly_sf %>% select(HOMIC_R,AREA,STFID)
plot(sf3)
```

> `sf` is an important package for data manipulation and integration with other programs (hence - conformance to formal ISO standards), it is probably not the most friendly tool for analytic visualization.  So ... we move on.


   

[^1]: For the correction of breaks after using classIntervals with spplot/levelplot see here http://r.789695.n4.nabble.com/SpatialPolygon-with-the-max-value-gets-no-color-assigned-in-spplot-function-when-using-quot-at-quot-r-td4654672.html

## 4. Choropleth mapping with `ggplot2`

[`ggplot2`](http://ggplot2.org/) is a widely used and powerful plotting library for R. It is not specifically geared towards mapping, but one can generate great maps. 

As we already have some reasonable experience with ggplot2, let me simply explain this way: we have made plots in ggplot2 using `geom_bar`, and `geom_line`, and ....  We are now simply going to use a new *geometry*.

If you read my "source" for this lesson (dtd May, 2017), you will see a significant explanation on how to use ggplot2 for mapping.  However, that approach is no longer required. 

--------

R continually evolves.  The advent of the `sf` package, which follows a `tidy` structuring of the sf objects allowed for an easy inclusion into the ggplot2 world.  A new geometry, `geom_sf` is available in the development branch of ggplot2.

You can install that version of ggplot2, and then the code below (using ggplot2) will work:  To install

    # Or the the development version from GitHub:
    # install.packages("devtools")
    devtools::install_github("tidyverse/ggplot2")


-----------

Here you see our Philly map, where I have used standard "ggplot" terminology, to include plotting with transparency.

```{r }
ggplot(philly_sf) +
  geom_sf(aes(fill=HOMIC_R),alpha=.5) + 
  scale_fill_viridis()+
  theme_minimal()
```

##### **(a)** Using ggplot2, create your own chloropleth plot using the Philadephia, but plotting a different variable from the dataset.

```{r}
ggplot(philly_sf) +
  geom_sf(aes(fill=TOTALPO),alpha=.5) + 
  scale_fill_viridis()+
  theme_minimal()
```

##### **(b)** Create another chloropleth plot using the Philadephia, but plotting a derived variable that would give number of Homicides per square mile.  (Note - the AREA variable is in ft^2)

```{r}
philly_sf$HOMIC_M<-philly_sf$N_HOMIC/philly_sf$AREA*3.59E-8
ggplot(philly_sf) +
  geom_sf(aes(fill=HOMIC_M),alpha=.5) + 
  scale_fill_viridis()+
  theme_minimal()
```



# STOP HERE

The material below requires you to register for a Google API key, which I am not asking you to do.

## 5. Basemaps with ggmap

`ggmap` builds on `ggplot` and allows to pull in tiled basemaps from different services, like Google Maps and OpenStreetMaps[^6].

In addition to providing a "base layer" for visualizations, ggmap also supports measuring and routing functionality, with routing as accessed via interfacing with the google routing api.

However, since our primary focus is visualization, (as opposed to analysis) lets focus on the visualization piece.

As key concept, ggmap allows you to put a map background at the foundation of a ggplot.  Then, you can add essesntially any type of layers to that plot using ggplot2 geometries, labels, etc.

So let's overlay the map from above on a google satellite basemap.

---

#### Exercise 3

First we use the `get_map()` command from `ggmap` to pull down the basemap. We need to tell it the location or the boundaries of the map, the zoom level, and what kind of map service we like (default is Google terrain). It will actually download the tile. `get_map()` returns a ggmap object, name it `ph_basemap`.  

    In order to view the map we then use `ggmap(ph_basemap)`.

    Look up the syntax of `?get_map()`, go back and forth between `get_map(..)` and `ggmap(ph_basemap)` to find the correct parameters for our example.


```{r }
#ph_basemap <- get_map(location="Philadelphia, PA", zoom=11, maptype = 'terrain',source="stamen")
#ggmap(ph_basemap)
```

Then we can reuse the code from the ggplot example above, replacing the first line, where we initialized a ggplot object above
        
        ggplot() + 
        ...

    with the line to call our basemap:

        ggmap(ph_basemap) +
        ...
     
    
    See if you can copy and paste this together. 




```{r}
# ph_3857 <- st_transform(philly_sf, 3857)
# 
# ggmap_bbox <- function(map) {
#   if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
#   # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
#   # and set the names to what sf::st_bbox expects:
#   map_bbox <- setNames(unlist(attr(map, "bb")), 
#                        c("ymin", "xmin", "ymax", "xmax"))
# 
#   # Coonvert the bbox to an sf polygon, transform it to 3857, 
#   # and convert back to a bbox (convoluted, but it works)
#   bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
# 
#   # Overwrite the bbox of the ggmap object with the transformed coordinates 
#   attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
#   attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
#   attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
#   attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
#   map
# }
# 
# ph_basemap <- ggmap_bbox(ph_basemap)
# 
# 
# ggmap(ph_basemap)+
#   coord_sf(crs = st_crs(3857)) + 
#   geom_sf(data=ph_3857,inherit.aes=FALSE,
#           aes(fill=HOMIC_R)) + 
#     scale_fill_viridis() 
  
```


#  More on this lesson will appear later this week.





